Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-09-15 draft_lm1.tex
Changing to PLOS template

\documentclass[12pt]{article}              
\usepackage[margin=1.5cm]{geometry}
\usepackage{setspace,relsize}               
\usepackage{moreverb}                        
\usepackage{url}
\usepackage{hyperref}
\hypersetup{colorlinks=true,citecolor=blue}
\usepackage{amsmath}
\usepackage{mathtools} 
\usepackage{amssymb}
\usepackage{indentfirst}
\usepackage[authoryear,round]{natbib}
\bibliographystyle{apalike}
\usepackage[pdftex]{lscape}
% \usepackage{float}
% \usepackage{longtable}
% \usetikzlibrary{arrows}
%%%%
\title{
Estimating the Initial Susceptible Fraction in Dengue Dynamics
}           
\author{Flavio Code\c{c}o Coelho \\
School of Applied Mathematics, Get\'ulio Vargas Foundation (FGV) \\
\& \\
Luiz Max de Carvalho \\
Program for Scientific Computing (PROCC), Oswaldo Cruz Foundation}

\date{\today}
% Notation defs
\def \rr {$R_{t}\ $}

\usepackage{Sweave}
\begin{document}                                  
\input{draft_lm1-concordance}

%
\maketitle
\begin{abstract}
We derive confidence/credibility bounds for the uncertainty about the estimate of the effective reproductive number $R_t$ under the assumption that reported disease counts are Poisson distributed.
It is shown that one can construct a conservative confidence interval for \rr using the induced probability distribution that results from the approach of \citet{mantel}.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Background}
\label{sec:intro}

In monitoring of infectious diseases, it is important to assess whether the incidence of a a particular disease is increasing significantly, in order to decide to take preventive measures.
The effective reproductive number at time $t$, \rr, can be understood as a real-time estimate of the basic reproductive number ($R_{0}$) and is defined as the average number of secondary cases per primary case at time $t$.
Let $Y_t$ be the number of reported disease cases for a particular time $t \in (0, T)$.
\citet{nishiura} extend the theory developed by \citet{stallybrass} and propose to estimate \rr as
\begin{equation}
\label{eq:Rtestimate}
R_t = \left( \frac{Y_{t+1}}{Y_t}\right)^{1/n}
\end{equation}
where $n$ is taken to be the ratio between the length of reporting interval and the mean generation time of the disease.
Here we are interested in the simpler case $n=1$.
If \rr is to be used as a decision tool, one needs to be able to quantify the uncertainty about estimate in equation~\ref{eq:Rtestimate}. 
Here we detail how to obtain confidence intervals for \rr, exploring two approaches under the assumption that the counts $Y_t$ are Poisson distributed for all $t$.

\newpage
\section{Approaches}

\subsection{Ederer \& Mantel (1974)}
\label{sec:mantel}

First we explore the approach of \citet{mantel}, whose objective is to obtain confidence intervals for the ratio of two Poisson counts. 
Let $Y_{t} \sim Poisson(\lambda_t)$ and $Y_{t+1} \sim Poisson(\lambda_{t+1})$ and define $S = Y_{t} + Y_{t+1}$.
The authors then proceed to note that by conditioning on the sum $S$
\begin{align}
\label{eq:binlike}
Y_{t+1} | S &\sim Binomial(S, \theta_t) \\
\theta_t &= \frac{\lambda_{t+1}}{\lambda_{t} + \lambda_{t+1}}
\end{align}
Let $c_{\alpha}(\theta_t) = \{\theta_t^{(L)} , \theta_t^{(U)} \}$ be such that $Pr(\theta_t^{(L)}<\theta_t <\theta_t^{(U)}) = \alpha$.
Analogously, define $c_{\alpha}(R_t) = \{R_t^{(L)} , R_t^{(U)} \}$ such that $Pr(R_t^{(L)}<R_t<R_t^{(U)}) = \alpha$.
\citet{mantel} show that one can construct a $100\alpha \%$ confidence interval for \rr by noting that
\begin{align}
 R_t^{(L)} &= \frac{\theta_t^{(L)}}{(1-\theta_t^{(L)})}\\
 R_t^{(U)} &= \frac{\theta_t^{(U)}}{(1-\theta_t^{(U)})}
\end{align}

Many authors have derived $c_{\alpha}(\theta_t)$ following orthodox approaches such as those of \citet{wilson} and \citet{clopper}, mainly for simplicity reasons.
We choose instead to take a Bayesian approach and use the  $100\alpha \%$ posterior credibility interval for $\theta_t$ as $c_{\alpha}(\theta_t)$.
If we choose the conjugate beta prior with parameters $a_0$ and $b_0$ for the binomial likelihood in (\ref{eq:binlike}), the posterior distribution for $\theta_t$ is
\begin{equation}
\label{eq:thetapost}
p(\theta_t| Y_{t+1}, S) \sim Beta(Y_{t+1} + a_0, Y_t + b_0)
\end{equation}
Equation~(\ref{eq:thetapost}) tells us that the induced distribution of $R_t$ is a beta prime with parameters $ a_1 = Y_{t+1} + a_0$ and $b_1 =  Y_t + b_0$.
The density of this induced distribution  $f_P(R_t)$ is then 
\begin{equation}
\label{eq:densityMantel}
f_P(R_t| a_1, b_1) = \frac{\Gamma(a_1 + b_1)}{\Gamma(a_1)\Gamma(b_1)} R_t^{a_1 - 1} (1 + R_t)^{-(a_1 + b_1)}
\end{equation}
Thus, the expectation of \rr under $f_P$ is $a_1/(b_1 - 1)$ and its variance is $a_1(a_1 + b_1 - 1)/\left((b_1 - 2)(b_1 - 1)^2 \right) $.

Since when $R_t > 1$ indicates sustained transmission, one may be interested in computing the probability of this event.
By noting that
\begin{align}
\label{cumprobMantel}
Pr(R_t > 1) &= 1 - \int_0^1 f_P(r)dr \\
            &= 1- Pr(\theta_t < \frac{1}{2})
\end{align}
one can avoid dealing with the density in~(\ref{eq:densityMantel}) directly.

% \subsection{Nishiura et al. (2010)}
% \label{sec:nishiura}
% 
% An alternative approach is to use the conditional distribution of \rr on $Y_{t+1}$ and $Y_t$ as defined in equation A7 of \citet{nishiura}:
% \begin{equation}
% \label{eq:unorm}
% f_{R}(R_{t}) = (Y_tR_{t})^{Y_{t+1}} e^{-Y_tR_{t}}
% \end{equation}
% Noticing the kernel of (\ref{eq:unorm}) is that of a gamma distribution with $a_2 = Y_{t+1}+1$ and $b_2 = Y_t$, we obtain a proper density from which to construct $c_{\alpha}(R_t)$, simply by computing the appropriate quantiles of said distribution.
%  This density is
% \begin{equation}
% \label{eq:densityNishiura}
% f_N(R_t| a_2, b_2) =  \frac{b_2^{a_2}}{\Gamma(a_2)} R_t^{a_2-1} e^{-b_2 R_t}
% \end{equation}
% Once again $Pr(R_t>1)$ can be easily computed using
% \begin{equation}
% \label{eq:cdfNishiura}
% P(R_t > 1) = 1-F_N(1) = 1- \frac{1}{\Gamma(a_2)} \gamma(a_2, b_2)
% \end{equation}
% where $F_N(R_t)$ is the distribution function of $R_t$ and $\gamma(\cdot)$ is the incomplete gamma function.

\subsection{A remark on prior distributions and tail behaviour}
\label{sec:tails}

In order to decide which approach to take, it may be of use analysing the variance and tail behaviour of the derived distributions for \rr. 
Consider the case of using a flat $Uniform(0, 1)$ prior for $\theta_t$ in Section~\ref{sec:mantel}.
With $a_0 = b_0 = 1$, $a_1 = a_2$ and $b_1 = b_2 + 1$.
The beta prime (inverse beta distribution) will have heavier tails compared to the conditional distribution in (\ref{eq:densityNishiura}), thus providing more conservative confidence/credibility intervals.  
As a side note, the Bayesian approach presented in Section~\ref{sec:mantel} will give similar results to those of \citet{wilson} and \citet{wilson} for $Y_{t+1}$ and $Y_t >> 1$.
Under the flat  uniform prior for $\theta_t$, the Bayesian posterior credibility interval is nearly indistinguishable from the confidence interval proposed by \citet{clopper} for $Y_{t+1}, Y_t > 20$.
Note that the $Beta(1, 1)$ uniform prior for $\theta_t$ constitutes a poor prior choice mainly because the induced distribution for \rr is only well-defined for $b_0 > 2$.

The main advange of the Bayesian approach is that one can devise prior distributions for $\theta_t$ taking advantage of the intuitive parametrisation and flexibility of the beta family of distributions.
Prior elicitation can also be done for \rr and the hyperparameters directly plugged into the prior for $\theta_t$. 
One can, for example, choose a priori mean and variance for \rr and find $a_0$ and $b_0$ that satisfy those conditions.
Let $m_0$ and $v_0$ be the prior expectation and variance for $R_t$. 
After some tedious algebra one finds
\begin{align}
\label{eq:elicitation}
a_0 &= \frac{m_0v_0 + m_0^3 + m_0^2}{v_0} \\
b_0 &= \frac{2v_0 + m_0^2 + m_0}{v_0}
\end{align}
If one wants only to specify $m_0$ and a coefficient of variation $c$ \footnote{$c = \sqrt{v_0}/ m_0$.} for $R_t$ \textit{a priori}, some less boring algebra gives:
\begin{align}
\label{eq:elicitationcv}
a_0 &= \frac{m_0^3c^2 + m_0^3 + m_0^2}{m_0^2c^2} \\
b_0 &= \frac{2m_0^2c^2 + m^2 + m}{m_0^2c^2}
\end{align}

\newpage
\section{Preliminary results -- Dengue notifications in Rio de Janeiro}
\label{sec:dataanalysis}

We now show the application of the methods described above to dengue notification data in the city of Rio de Janeiro.
The data consists of weekly dengue notifications for 10 Health Planning Areas (APS, in Portuguese) for the period 2010-2014.
Figure~\ref{fig:casesperAPS} shows the data.
We compute $c_{\alpha}(R_t)$ for the total (sum) number of notifications using the two approaches described, using a generation time of $3$ weeks and choose $a_0$ and $b_0$ such that $E[R_t] = 1/2$ and $Var(R_t) = 2$.
The resulting estimates and $95\%$ confidence bounds are presented in Figure~\ref{fig:confidencetotals}.
Finally, we compute $Pr(R_t > 1)$ using the two distributions discussed here (Figure~\ref{fig:pRt1}).

\section{Final Remarks}
\label{sec:remarks}

Both strategies presented here seem appropriate to the quantification of uncertainty about $R_t$.
The distribution obtained by combining the approach of \citet{mantel} with Bayesian inference for $\theta_t$, however, possesses heavier tails and thus offers more conservative bounds.
Additionally, the ability to specify prior information on $R_t$ based on expert knowledge may prove useful.

\section{Acknowlegements}
LMC is grateful to Dr. Leonardo Bastos for useful discussions.
\newpage
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\setkeys{Gin}{scale=.7}
\begin{center}
\includegraphics{FIGURES/figure-002}
\end{center}
\caption{Total number of dengue notifications per HPA (APS)}
\label{fig:casesperAPS}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\setkeys{Gin}{scale=.7}
\begin{center}
\includegraphics{FIGURES/figure-003}
\end{center}
\caption{\rr estimates for the total notifications and confidence intervals obtained using the two proposed methods.}
\label{fig:confidencetotals}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!h]
\setkeys{Gin}{scale=.7}
\begin{center}
\includegraphics{FIGURES/figure-004}
\end{center}
\caption{Probability that $R_t \geq 1$.
The interrupted gray line marks the arbitrary value of $0.80$.  
}
\label{fig:pRt1}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newpage
\bibliography{lm1}
\end{document}

>>>>

% Template for PLoS
% Version 2.0 July 2014
%
% To compile to pdf, run:
% latex plos.template
% bibtex plos.template
% latex plos.template
% latex plos.template
% dvipdf plos.template
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% Be advised that this is merely a template 
% designed to facilitate accurate translation of manuscript content 
% into our production files. 
%
% This template contains extensive comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template 
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication and enters production, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file and leave only
% the final text of your manuscript.
%
% DO NOT ADD EXTRA PACKAGES TO THIS TEMPLATE unless absolutely necessary.
% Packages included in this template are intentionally
% limited and basic in order to reduce the possibility
% of issues during our production process.
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% See http://www.plosone.org/static/figureGuidelines for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - tabs/spacing/line breaks within cells to alter layout
% - vertically-merged cells (no tabular environments within tabular environments, do not use \multirow)
% - colors, shading, or graphic objects
% See http://www.plosone.org/static/figureGuidelines#tables for table guidelines.
%
% For sideways tables, use the {rotating} package and use \begin{sidewaystable} instead of \begin{table} in the appropriate section. PLOS guidelines do not accomodate sideways figures.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://www.plosone.org/static/latexGuidelines
%
% Please be sure to include all portions of an equation in the math environment, and for any superscripts or subscripts also include the base number/text. For example, use $mathrm{mm}^2$ instead of mm$^2$ (do not use \textsuperscript command).
%
% DO NOT USE the \rm command to render mathmode characters in roman font, instead use $\mathrm{}$
% For bolding characters in mathmode, please use $\mathbf{}$ 
%
% Please add line breaks to long equations when possible in order to fit our 2-column layout. 
%
% For inline equations, please do not include punctuation within the math environment unless this is part of the equation.
%
% For spaces within the math environment please use the \; or \: commands, even within \text{} (do not use smaller spacing as this does not convert well).
%
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt]{article}
% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}
% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}
\usepackage{hyperref}
% line numbers
\usepackage{lineno}
% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }
% rotating package for sideways tables
%\usepackage{rotating}
% If you wish to include algorithms, please use one of the packages below. Also, please see the algorithm section of our LaTeX guidelines (http://www.plosone.org/static/latexGuidelines) for important information about required formatting.
%\usepackage{algorithmic}
%\usepackage{algorithmicx}
% Use doublespacing - comment out for single spacing
%\usepackage{setspace} 
%\doublespacing
% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm
% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}
% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}
% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}

%% Include all macros below. Please limit the use of macros.

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Estimating the Initial Susceptible Fraction in Dengue Dynamics}
}
% Insert Author names, affiliations and corresponding author email.
\\
Flavio Code\c{c}o Coelho$^{1}$, 
Luiz Max de Carvalho$^{2}$, 
Author3$^{3,\ast}$
\\
\bf{1} School of Applied Mathematics, Get\'ulio Vargas Foundation (FGV), Rio de Janeiro -- RJ, Brazil.
\\
\bf{2} Program for Scientific Computing (PROCC), Oswaldo Cruz Foundation, Rio de Janeiro -- RJ, Brazil.
\\
$\ast$ E-mail: fccoelho@fgv.br
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author Summary}

\section*{Introduction}
% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Methods}
\subsection*{The model}
\subsection*{Parameter estimation using Bayesian melding}
% Results and Discussion can be combined.
\section*{Results}

% We only support three levels of headings, please do not create a heading level below \subsubsection.
\subsection*{Subsection 1}

\section*{Discussion}

% Do NOT remove this, even if you are not including acknowledgments.

\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions.

\section*{References}
\bibliography{lm1}
\section*{Figure Legends}
% This section is for figure legends only, do not include
% graphics in your manuscript file.
%
%\begin{figure}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure caption.  
%}
%\label{Figure_label}
%\end{figure}

\section*{Tables}
% 
% See introductory notes if you wish to include sideways tables.
%
% NOTE: Please look over our table guidelines at http://www.plosone.org/static/figureGuidelines#tables to make sure that your tables meet our requirements. Certain types of spacing, cell merging, and other formatting tricks may have unintended results and will be returned for revision.
%
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}

\section*{Supporting Information Legends}
%
% Please enter your Supporting Information captions below in the following format:
%\item{\bf Figure SX. Enter mandatory title here.} Enter optional descriptive information here.
% 
%\begin{description}
%\item {\bf}
%\item {\bf}
%\end{description}

\end{document}
